Bad forecast, good decision: what makes a model relevant for determining policy

### 2020-10-16 manuscript/manuscript.Rmd
:pen:

---
title: "Bad forecast, good decision: accuracy can be a poor metric for policy"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
#    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
    With the rapid expansion of available data, it is frequently suggested that _forecasting performance_ is the ultimate test of a scientific model [e.g. @Dietz2018].  Yet the model that makes the most accurate forecasts is not always the model that yields the best decisions.  This is no accident of statistics, even formal criteria such as strictly proper scoring rules [@Gneiting2007] cannot determine the best model for decision-making because they do not consider the context of the decision process itself.  I illustrate this paradox using a classic example of ecological decision making in fisheries management.  This example also reveals how an understanding of the decision problem can both explain the poor management outcomes of model which produces the best forecast, and also see how to avoid making such a mistake.  As forecasting becomes an increasing attractive and viable possibility for assessing ecological models and applying them to real-world decisions, I hope this illustration serves as a reminder that modeling can not be done in a vacuum, but must reflect the context in which the model will be applied.
journal: Theoretical ecology
layout: 3p
bibliography: refs.bib
output: rticles::elsevier_article
keywords: 
  - transients, 
  - optimal control, 
  - adaptive management, 
  - stochasticity, 
  - uncertainty, 
  - ecological management
preamble: |
  \journal{preprint}
  ## EditorialManager adds its own linenumbers to pdf already
  \usepackage{lineno}
  \linenumbers

@@

This paradox in performance of forecasting vs performance in decision making can be easily resolved by considering the context of the decision problem more closely. Starting back in the origins of modern fisheries management in the 1950s, @Schaefer1954 demonstrates that the Maximum Sustainable Yield (MSY) is maintained by harvesting a stock down to the size at which it achieves its maximum growth rate, i.e. $K/2$ for a logistic model (or any other symmetric growth model).  The details of this optimal harvest have been refined to allow for dynamic optimization with discounted future profits [@Clark1974] stochastic population growth [@Reed1979], and even imperfect observations [@Memarzadeh2018], but Schaefer's observation is sufficient in this case to see that the decision problem depends very much on the value of $K$ while being essentially independent of the other model parameters, including the growth rate, $r$ and stochasticity, $\sigma$. However, the growth rate (and magnitude of the stochasticity) matter very much to forecast accuracy.  Figure 3 shows the functional form of our two logistic-curve models, compared to the functional form of the "true" model used to drive the simulations.  From this graph, it is clear to see that Model 2 does indeed lie closer to the true model throughout the state space, and agrees precisely with the carrying capacity (where both functions cross zero with negative slope).  However, because the model 3, the true model, is not symmetric, it reaches its maximum growth rate very near the stock size that also maximizes the growth rate of model 1, which is considerably higher stock size than that of model 2.  Consequently, model 2 leads to over-fishing, while model 1 leads to nearly-optimal management, despite its poor performance as a forecaster.

If we had access to model 3, we would no doubt find that it outperformed model 2 in forecast accuracy as well as ecological and economic performance.  But in real ecological decision making, we never know the true model -- we will always be comparing among approximations.

Within fisheries, even in age-structured models, recruitment approximations with symmetric growth functions (logistic, Ricker, Beverton-Holt, etc) still dominate.

>>>>

---
title: "Bad forecast, good decision: what makes a model relevant for determining policy"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
#    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
    With the rapid expansion of available data, it is frequently suggested that _forecasting performance_ is the ultimate test of a scientific model [e.g. @Dietz2018].  Yet the model that makes the most accurate forecasts is not always the model that yields the best decisions.  This is no accident of statistics, even formal criteria such as strictly proper scoring rules [@Gneiting2007] cannot determine the best model for decision-making because they do not consider the context of the decision process itself.  I illustrate this paradox using a classic example of ecological decision making in fisheries management.  This example also reveals how an understanding of the decision problem can both explain the poor management outcomes of model which produces the best forecast, and also see how to avoid making such a mistake.  As forecasting becomes an increasing attractive and viable possibility for assessing ecological models and applying them to real-world decisions, I hope this illustration serves as a reminder that modeling can not be done in a vacuum, but must reflect the context in which the model will be applied.
journal: Theoretical ecology
layout: 3p
bibliography: refs.bib
output: rticles::elsevier_article
keywords: 
  - transients, 
  - optimal control, 
  - adaptive management, 
  - stochasticity, 
  - uncertainty, 
  - ecological management
preamble: |
  \journal{preprint}
  ## EditorialManager adds its own linenumbers to pdf already
  \usepackage{lineno}
  \linenumbers

@@

This paradox in performance of forecasting vs performance in decision making can be easily resolved by considering the context of the decision problem more closely. Starting back in the origins of modern fisheries management in the 1950s, @Schaefer1954 demonstrates that the Maximum Sustainable Yield (MSY) is maintained by harvesting a stock down to the size at which it achieves its maximum growth rate, i.e. $K/2$ for a logistic model (or any other symmetric growth model).  The details of this optimal harvest have been refined to allow for dynamic optimization with discounted future profits [@Clark1974] stochastic population growth [@Reed1979], and even imperfect observations [@Memarzadeh2018], but Schaefer's observation is sufficient in this case to see that the decision problem depends very much on the value of $K$ while being essentially independent of the other model parameters, including the growth rate, $r$ and stochasticity, $\sigma$. However, the growth rate (and magnitude of the stochasticity) matter very much to forecast accuracy.  Figure 3 shows the functional form of our two logistic-curve models, compared to the functional form of the "true" model used to drive the simulations.  From this graph, it is clear to see that Model 2 does indeed lie closer to the true model throughout the state space, and agrees precisely with the carrying capacity (where both functions cross zero with negative slope).  However, the peak of model 3 very nearly matches the peak of model 1, and falls higher stock size than that of model 2.

Thus, each year our model 1 managers are again chagrined to see the stock size estimates come in far below their rosy predictions, but nevertheless manage to set a nearly optimal quota by comparing the observed stock size to the model's predicted optimal escapement level [@Reed1979].  Meanwhile, model 2 managers could only congratulate themselves that each year's observations fall neatly within their predicted interval, unaware that the they were over-exploiting the fishery by both economic and ecological metrics.  If we had access to model 3, we would no doubt find that it outperformed model 2 in forecast accuracy as well as ecological and economic performance.  But in real ecological decision making, we never know the true model -- we will always be comparing among approximations.  Within fisheries, even in today's parameter-rich age-structured models, recruitment approximations with symmetric growth functions (Logistic, Ricker, Beverton-Holt, etc) still dominate [@ramlegacy].

This issue is by no means unique to fisheries.  Throughout resource management and conservation decision-making, as is no doubt true in other fields, decisions about which model to use are guided by which model best fits the data []. As this example illustrates, such metrics, no matter how rigorously defined, may select entirely the wrong model.  And once we are managing with the wrong model, no amount of comparing model predictions to outcomes will guarantee we discover our mistake.  Only by comparing to the outcomes we actually care about -- in this case, economic and ecological performance -- can we see which model is best for decision-making.  Conversely, our example reminds us that just because a model continually makes poor predictions does not mean that the model leads to poor decisions.  Model 1, despite it's many mistakes, is right about one key feature: the position of the peak growth rate -- and that is enough to guarentee nearly optimal performance.   This should also be reassuring to modelers and decision makers, for it reminds us that effective models need be correct or even close to correct in every aspect, as long as they capture the key features of the decision context.  Research into decision theory, such as the work by @Schaefer1954; @Clark1974; @Reed1979; @Memarzadeh2018 help us better understand that context.  Adaptive management approaches [@Walters1978] can apply that theory to compare management outcomes between models directly, rather than comparing to forecast performance.
